A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-07 latex/main.tex
Updates from Overleaf

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024},
and provide a natural and efficient means of 
systematically incorporating recombination 
into the study of SARS-CoV-2.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019,Gopalan2025bridging}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2025robust} have led to a surge 
of interest in their application in population and statistical 
genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
% This is necessary because a bunch of the phylo literature refers
% to ARGs, but it's all in the "coalescent with recombination" sense
% Rannala2025 dismisses ARG-based methods at a stroke because they're
% all based on the assumption of a single panmictic population 
A subtle but important point distinguishing recent work on ARGs
from the classical literature is that the ARG 
``data structure''\cite{Minichiello2006mapping,Wong2024} is no longer
tightly coupled to the generative
process\cite{Hudson1983,Griffiths1997ancestral}, substantially 
broadening the range of potential applications~\cite{Bisschop2025likelihoods}.
Another important point is that strong software support for ARGs now
exists, and the tskit library~\cite{Ralph2020} is the basis of a 
rapidly growing 
software ecosystem of 
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
tools.

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Evidence for recombination in the ARG.
All plots show ``robust'' sc2ts recombination nodes,
colored according to their classification by rebar
(recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table~\ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
Jackson group B (JB) and Jackson singleton recombinant 1 (J1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
leading to XA, XW, XBB, and an
undesignated recombination event under XBB, which we here refer to as  XBB.x.}
\label{fig:averted_mutations}
\end{figure}

@@

We explored the spatial distribution of the breakpoint locations of
the ``robust'' recombination events along the genome,
and quantified how precisely breakpoints were localized.
For each event, we computed an interval estimate of its breakpoint location,
using its size as a measure of precision (median, 2,207 bases; STAR Methods).

>>>>

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024},
and provide a natural and efficient means of 
systematically incorporating recombination 
into the study of SARS-CoV-2.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019,Gopalan2025bridging}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2025robust} have led to a surge 
of interest in their application in population and statistical 
genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
% This is necessary because a bunch of the phylo literature refers
% to ARGs, but it's all in the "coalescent with recombination" sense
% Rannala2025 dismisses ARG-based methods at a stroke because they're
% all based on the assumption of a single panmictic population 
A subtle but important point distinguishing recent work on ARGs
from the classical literature is that the ARG 
``data structure''\cite{Minichiello2006mapping,Wong2024} is no longer
tightly coupled to the generative
process\cite{Hudson1983,Griffiths1997ancestral}, substantially 
broadening the range of potential applications~\cite{Bisschop2025likelihoods}.
Another important point is that strong software support for ARGs now
exists, and the tskit library~\cite{Ralph2020} is the basis of a 
rapidly growing software ecosystem of
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualization~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
tools.

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Evidence for recombination in the ARG.
All plots show QC-passing sc2ts recombination nodes,
colored according to their classification by rebar
(recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table~\ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
Jackson group B (JB) and Jackson singleton recombinant 1 (J1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each QC-passing recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
leading to XA, XW, XBB, and an
undesignated recombination event under XBB, which we here refer to as  XBB.x.}
\label{fig:averted_mutations}
\end{figure}

@@

We explored the spatial distribution of the breakpoint locations of
the QC-passing recombination events along the genome,
and quantified how precisely breakpoints were localized.
For each event, we computed an interval estimate of its breakpoint location,
using its size as a measure of precision (median, 2,207 bases; STAR Methods).
